A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/DeadlockFreedom.lagda.tex
revised deadlock freedom

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalize in Agda. First of all we need to define the notion of
\emph{thread}, that is any process other than a cut. As it turns out, it is
convenient to provide an more fine-grained classification of threads,
distinguishing between links, input/output actions, and sometimes also whether
such actions operate on free or bound channels. We define predicates for each of
these process classes:

@@

These predicates are relatively uninteresting since they are essentially
isomorphic to the relevant parts of the \AgdaDatatype{Proc} datatype. The only
relevant aspect that is worth pointing out here is that in \AgdaDatatype{Input},
\AgdaDatatype{Output} and \AgdaDatatype{Server} the channel being acted upon by
the thread is the \emph{first} in $\Context$, hence it is the \emph{most
recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
those threads that, in the context of a cut, operate on the channel bound by the
cut or on a free channel. As an illustration, consider the
\AgdaInductiveConstructor{wait} case in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

@@

\begin{code}
  wait     : ∀{C Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Delayed (wait (ch ⟨ >_ {C} p ⟩ P))
\end{code}
where the use of the \AgdaInductiveConstructor{>} constructor indicates that the
thread operates on a channel other than the most recent one. Note that here we
have to specify the type $C$ in front of $\Context$ or else Agda is unable to
resolve some metavariables.

@@

We can now define observability, reducibility and aliveness in the expected way:

@@

When $P$ is a thread, then it is obviously observable and hence alive. When $P$
is a cut $\Cut[A]\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied
recursively on $P$ and $Q$. If either of these applications yields a reduction,
then the whole cut is reducible and therefore alive. If both applications yield
a thread, then we conclude that the cut is alive first rewriting it into a
canonical cut (\AgdaFunction{canonical-cut}) and then using
\AgdaFunction{canonical-cut-alive}.

>>>>

As we have seen in \Cref{sec:calculus}, the deadlock freedom result
(\Cref{thm:df}) rests on some notions and predicates about processes which we
now formalize in Agda. First of all we need to define the notion of
\emph{thread}, that is any process other than a cut. As it turns out, it is
convenient to provide a more fine-grained classification of threads,
distinguishing between links, input/output actions, and sometimes also whether
such actions operate on free or bound channels. We define predicates for each of
these process classes:

@@

These predicates are relatively uninteresting since they are essentially
isomorphic to the relevant fragments of the \AgdaDatatype{Proc} datatype. The
only aspect that is worth pointing out here is that in \AgdaDatatype{Input},
\AgdaDatatype{Output} and \AgdaDatatype{Server} the channel being acted upon by
the thread is the \emph{first} in $\Context$, hence it is the \emph{most
recently introduced} channel, whereas in \AgdaDatatype{Delayed} and
\AgdaDatatype{DelayedServer} the channel is older. This allows us to distinguish
those threads that, in the context of a cut, operate on the channel bound by the
cut or on a free channel. To clarify this aspect, consider the
\AgdaInductiveConstructor{wait} constructor in \AgdaDatatype{Input} and in
\AgdaDatatype{Delayed}. In the former predicate we have

@@

\begin{code}
  wait     : ∀{C Γ Δ P} (p : Γ ≃ [ ⊥ ] + Δ) → Delayed (wait (ch ⟨ >_ {C} p ⟩ P))
\end{code}
where the use of the \AgdaInductiveConstructor{>} constructor indicates that the
thread operates on a channel other than the most recent one. Note that here we
have to specify the type $C$ in front of $\Context$ or else Agda is unable to
automatically resolve some metavariables.

@@

We define observability, reducibility and aliveness in the expected way:

@@

When $P$ is a thread, then it is obviously observable and hence alive. When $P$
is a cut $\Cut[A]\x{P}{Q}$, \AgdaFunction{deadlock-freedom} is applied
recursively on $P$ and $Q$. If either of these applications yields a reduction,
then the whole cut is reducible and therefore alive. If both applications yield
a thread, then we conclude that the cut is alive first rewriting it into a
canonical cut with \AgdaFunction{canonical-cut} and then using
\AgdaFunction{canonical-cut-alive}.
